Sampling Bayesian probabilities given only sampled priors

### 2024-11-19 main.tex
typo

% section/paper on sampling with sampled priors.
%\documentclass[preprint,linenumbers]{aastex63}
\documentclass[linenumbers, onecolumn]{aastex631}
\usepackage{xcolor}

@@

\newcommand{\ie}{\textit{i.e.}}
\newcommand{\eg}{\textit{e.g.}}
\newcommand{\E}{\mathrm{E}}
\newcommand{\eqq}[1]{Equation~(\ref{#1})}
\newcommand\gary[1]{{\color{red} \{\textbf{GMB}: #1\}}}
%\tabletypesize{\footnotesize}
\newcommand{\vecc}{\ensuremath{\mathbf{c}}}
\newcommand{\vecq}{\ensuremath{\mathbf{q}}}
\newcommand{\vecn}{\ensuremath{\mathbf{n}}}
\newcommand{\vecu}{\ensuremath{\mathbf{u}}}
\newcommand{\hatc}{\ensuremath{\hat{\mathbf{c}}}}
\newcommand{\vecP}{\ensuremath{\mathbf{P}}}
\newcommand{\likeli}{\mathcal{L}}

\begin{document}

@@

\section{Introduction} \label{sec:intro}

@@

A more rigorous and extremely efficient method of marginalizing over high-dimensional nuisance parameters was proposed by \citet{hans}, for the case where the following restrictions apply:
\begin{enumerate}
\item The likelihood of the observable \vecc\ is normal, $\vecc \sim \mathcal{N}( \hatc, \Sigma_c).$
\item The prior $p(\vecn)$ can also be assumed to be normal, with a mean and covariance matrix $\Sigma_n$ that in our case could be assigned from the mean and covariance of the samples of \vecn\ we are given.
\item The model \hatc\ can be linearized about fiducial values $\vecq_0, \vecn_0$ without loss of accuracy exceeding measurement errors.
\end{enumerate}
Under these conditions, \citet{hans} show that the marginalization over \vecn\ is equivalent to adding terms to $\Sigma_c,$ such that any MC process need not sample \vecn\ at all.

@@

@@

\end{acknowledgments}
% \bibliography{references}
\bibliographystyle{aasjournal}
\bibitem[Cordero et al.(2022)]{hyperrank} Cordero, J.~P., Harrison, I., Rollins, R.~P., et al.\ 2022, \mnras, 511, 2170. doi:10.1093/mnras/stac147

\bibitem[Hadzhiyska et al.(2020)]{hans} Hadzhiyska, B., Alonso, D., Nicola, A., et al.\ 2020, \jcap, 2020, 056. doi:10.1088/1475-7516/2020/10/056

>>>>

% section/paper on sampling with sampled priors.
% \documentclass[preprint,linenumbers]{aastex631} # 6.31 fails with \align!!
\documentclass[linenumbers, onecolumn]{aastex63}
\usepackage{xcolor}

@@

\newcommand{\ie}{\textit{i.e.}}
\newcommand{\eg}{\textit{e.g.}}
\newcommand{\E}{\mathrm{E}}
\newcommand{\eqq}[1]{Equation~(\ref{#1})}
\newcommand\gary[1]{{\color{red} \{\textbf{GMB}: #1\}}}
%\tabletypesize{\footnotesize}
\newcommand{\vecc}{\ensuremath{\mathbf{c}}}
\newcommand{\vecq}{\ensuremath{\mathbf{q}}}
\newcommand{\vecn}{\ensuremath{\mathbf{n}}}
\newcommand{\vecu}{\ensuremath{\mathbf{u}}}
\newcommand{\hatc}{\ensuremath{\hat{\mathbf{c}}}}
\newcommand{\vecP}{\ensuremath{\mathbf{P}}}
\newcommand{\covm}{C}
\newcommand{\matA}{\ensuremath{A}}
\newcommand{\matD}{D}
\newcommand{\matU}{U}
\newcommand{\matV}{V}
\newcommand{\matW}{W}
\newcommand{\proj}{P}  % Projection matrix
\newcommand{\jac}{J}   % Jacobian
\newcommand{\ident}{I}  % identity
\newcommand{\DD}{\Delta_D}
\newcommand{\likeli}{\mathcal{L}}
\newcommand{\trace}{\text{Tr}}
\begin{document}

@@

\section{Motivation} \label{sec:intro}

@@

A more rigorous and extremely efficient method of marginalizing over high-dimensional nuisance parameters was proposed by \citet{hans}, for the case where the following restrictions apply:
\begin{enumerate}
\item The likelihood of the observable \vecc\ is normal, $\vecc \sim \mathcal{N}( \hatc, \covm_c),$ with $\covm_c$ fixed.
\item The prior $p(\vecn)$ can also be assumed to be normal, with a mean and covariance matrix $\covm_n$ that in our case could be assigned from the mean and covariance of the samples of \vecn\ we are given.
\item The model \hatc\ can be linearized in \vecn\ about fiducial values $\vecq_0, \vecn_0$ without loss of accuracy exceeding measurement errors, with the derivatives independent of \vecq.
\end{enumerate}
Under these conditions, \citet{hans} show that the marginalization over \vecn\ is equivalent to adding terms to $\covm_c,$ such that any MC process need not sample \vecn\ at all.

@@

\section{Derivation}\label{sec:deriv}
We assume that we do have a multivariate normal likelihood for the
observables \vecc\, with the mean being some model
$\hatc(\vecq,\vecn)$ and a fixed covariance matrix $\covm_c.$ In this
section we will assume that the \vecn\ vectors have been shifted by
the mean $\vecn_0 \equiv \left\langle \vecn_\alpha \right\rangle$ so
that their mean is zero.
We seek some function $\hat{\vecn}(\vecu)$ of a lower-dimensional vector \vecu\ which can be substituted for \vecn\ and yield nearly the same likelihood function for any \vecn\ in the domain spanned by the samples $\vecn_\alpha,\; \alpha\in 1\ldots N_\alpha,$ that we are provided.  This means we want a map $\vecn_\alpha\rightarrow \vecu_\alpha$ and a function $\hat{\vecn}(\vecu)$ which minimize the error in log-likelihood
\begin{equation} -2 \left\langle \log \likeli(\vecc | \vecq,\vecn) - \log \likeli(\vecc | \vecq,\hat{\vecn}(\vecu)) \right\rangle =
  \left\langle \Delta\chi^2 \right\rangle \approx \frac{1}{N_\alpha} \sum_\alpha
                                            \left[ \hatc(\vecq,\vecn_\alpha) - \hatc(\vecq,\hat{\vecn}(\vecu_\alpha)) \right]^T
                                            \covm_c^{-1}
                                            \left[ \hatc(\vecq,\vecn_\alpha) - \hatc(\vecq,\hat{\vecn}(\vecu_\alpha)) \right].
                                          \end{equation}

If we next define the Jacobian matrix
\begin{equation}
  \jac \equiv \left.\frac{\partial\hatc}{\partial\vecn}\right|_{\vecq_0, \vecn_0}
\label{eq:jacobian}
\end{equation}
and assume that a first-order Taylor expansion about the nominal values is valid across the domain of interest, then we obtain
\begin{equation}
  \left\langle \Delta\chi^2 \right\rangle \approx \frac{1}{N_\alpha} \sum_\alpha
  \left[ \vecn_\alpha - \hat{\vecn}(u_\alpha)\right]^T \jac^T \covm_c^{-1} \jac
  \left[ \vecn_\alpha - \hat{\vecn}(u_\alpha)\right].
  \label{eq:linearized}
\end{equation}
We ask what is the optimal \emph{linear} model $\hat{\vecn} = \matU \cdot \vecu$ to use.
It is useful to define the matrix \matD\ and its ``square root'' $\DD$ by
\begin{equation}
   \matD \equiv \jac^T \covm_c^{-1} \jac =  \DD^T \DD.
 \end{equation}
Since \matD\ is positive-definite and symmetric, $\DD$ will exist and be real-valued.
If \vecn\ has $N$ elements, and we want to compress the representation
to a vector \vecu\ with $M$ elements, we will use the $M\times N$ projection
matrix $\proj_M$ whose elements are
\begin{equation}
  \left(\proj_M\right)_{ij} \equiv
\begin{cases}
                                            1,  &  i=j\le M \\
                                            0,  & \text{otherwise.}
\end{cases}
\end{equation}
such that $\proj_M \proj_M^T=\ident_M,$ the $M\times M$ identity
matrix. We will assume that \matU\ satisfies
\begin{equation}
  \DD \matU = \matV_U \proj_M^T
\label{eq:vupm}
\end{equation}
for some unitary matrix $\matV_U$.  We can do this because, for any $\matU$, we have the
singular-value decomposition $\DD \matU = \matV_U P_M^T \Lambda_U
W_U^T,$ where $\Lambda_U$ is diagonal, $W_U$ is unitary, and both are
$M\times M$.  We can always absorb these last two factors into the \vecu\ values
to conserve the values of $\hat{\vecn} = \matU \vecu$ while making
\eqq{eq:vupm} hold.

With these definitions and a choice of \matU, the $\chi^2$ error and the values of $\vecu_\alpha$ that minimize it become
\begin{align}
  \left\langle \Delta\chi^2 \right\rangle & \approx \frac{1}{N_\alpha} \sum_\alpha
                                            \left[ \vecn_\alpha - \matU \vecu_\alpha\right]^T \matD^{-1}
                                            \left[ \vecn_\alpha - \matU \vecu_\alpha\right] \\
  \Rightarrow \quad \vecu_{\alpha,\text{min}} & = \left[ \matD -
                                                \matD\matU\left(\matU^{T}\matD\matU\right)^{-1}
                                                \matU^T\matD\right]\vecn_\alpha  \\
                                          & = \proj_M \matV_U^T \DD \vecn_\alpha.
\end{align}                                            
Defining the covariance matrix of the samples $\vecn_\alpha$ as
$\covm_n=\left\langle \vecn_\alpha \vecn^T_\alpha \right\rangle,$ and
plugging the last line back into the first,  gives us the quantity we
want to minimize by selecting \matU\ properly:
\begin{align}
  \left\langle \Delta \chi^2 \right\rangle 
  & = \trace\left[\covm_n\matD
    - \covm_n \DD^T \matV_U \proj_M^T \proj_M \matV_U^T \DD \right], \\
 & = \trace\left[\matV_A \Lambda_A \matV_A^T 
    - \Lambda_A  \matV_A^T \matV_U \proj_M^T \proj_M \matV_U^T
   \matV_A\right], \\
  & = \sum_{i=1}^N \lambda^A_i \left[ 1 - \sum_{j=1}^M (\matV^T_A
    \matV_U)^2_{ij}\right], \label{eq:dchi2} \\
  \text{with} \quad \matA & \equiv \DD \covm_c \DD^T = \matV_A \Lambda_A \matV_A^T.
\end{align}
The last line defines a new symmetric, non-negative matrix \matA\ that
has eigenvectors $\matV_A$ and positive eigenvalues $\Lambda_A=\text{Diag}(\lambda^A_1,
\lambda^A_2, \ldots, \lambda^A_N).$ We assume the eigenvalues are
placed into decreasing order.

The minimization of $\left\langle \chi^2 \right\rangle$ thus requires
choosing $\matV_U,$ to minimize \eqq{eq:dchi2}, then retrieving \matU\
from \eqq{eq:vupm}.  Since $\matV_A^T\matV_U$ is a unitary matrix,
each term in brackets must be non-negative and $\le 1,$ and their total must equal
$N-M$.  The optimal case would be to have the bracketed term be zero
for $i\le M$, and be unity when multiplying the smaller $\lambda^A_i$
at $i>M.$  This occurs when $\matV_U=\matV_A,$ which is therefore our
optimal compression in terms of reproducing the likelihood of the
data.

The desired linear compression of \vecn\ therefore has the following properties:
\begin{itemize}
\item The mean error in $2\log \likeli$ is
  \begin{equation}
    \left\langle \Delta\chi^2 \right \rangle = \sum_{i=M+1}^N
    \lambda^A_i,
  \end{equation}
  and we will want to choose the dimension $M$ of the compression to
  keep this quantity, the observable effect of the parts of \vecn\ that we 
  project away,  below unity.
  Note that the $\lambda^A_i$ values are determined by a combination of
  the covariance matrix of the samples of \vecn, along with the Jacobian
  matrix and the covariance $\covm_c$ of the observables, letting us
  know the measurement significance of each mode of variation of \vecn.

  \item The \matU\ matrix containing the $M$ basis vectors for
    reconstructing \vecn\ is
    \begin{equation}
      \matU = \DD^{-1} \matV_A \proj_M^T.
    \end{equation}
    This requires that $\DD$ be invertible, but if we recall that
    $\matV_A$ are the eigenvectors of $\DD \covm_n \DD,$ we can see
    that we can get a well-defined result as long as the rank of
    \matD\ is $\ge M$.  ???

  \item The compressed $\vecu_\alpha$ for a chosen sample of \vecn\ is
    \begin{equation}
      \vecu_\alpha = \proj_M \matV_A^T \DD \vecn_\alpha,
    \end{equation}
    and the covariance matrix of the compressed samples is
    \begin{equation}
      \covm_u = \left\langle \vecu \vecu^T \right\rangle = \proj_M
      \matV_A^T \DD \covm_n \DD^T \proj_M^T = \text{Diag}(\lambda^A_1,
      \lambda^A_2, \ldots \lambda^A_M).
    \end{equation}
The components of the compressed \vecu\ coefficients are therefore
uncorrelated.  Component $u_i$ has a variance
$\lambda^A_i,$ which is also equal to the reduction in $\left\langle
  \Delta \chi^2 \right\rangle$ that we will accrue if we include basis
vector $\mathbf{U}_i$ into our linear reconstruction of \vecn.
\end{itemize}

Upon executing the compression $\{\vecn_\alpha\} \rightarrow
\{\vecu_\alpha\},$ we may find the distribution of the samples to be
consistent with a multivariate normal---which will have the diagonal covariance
matrix $\covm_u,$ and statistical independence for the elements of
\vecu.
There is,however,  no \textit{a prior} reason that
this must be the case.  We can use any density estimator over
$\{\vecu_\alpha\}$ to define a continuous $p(\vecu)$ for a Markov
chain running over the $(\vecq,\vecu)$ space.  This is the principle
advantage of our method over the single-step covariance-inflation
method of \citet{hans}.

Even if the \vecu\ prior is well approximated as multivariate normal,
there are practical advantages of compressing the nuisance 
variables and retaining them in the cosmological Markov chain rather
than using the covariance-inflation method for analytic
marginalization.  One can, for example, examine the posterior
distributions of \vecu\ and see if they are at the edges of the prior,
which would potentially indicate an inconsistency between the data and
the prior.

\section{Application}\label{sec:app}

xxxxxxxxx xxxxxxxxx xxxxxxxxx xxxxxxxxx xxxxxxxxx xxxxxxxxx 
xxxxxxxxx xxxxxxxxx xxxxxxxxx xxxxxxxxx xxxxxxxxx xxxxxxxxx 
xxxxxxxxx xxxxxxxxx xxxxxxxxx xxxxxxxxx xxxxxxxxx xxxxxxxxx 
xxxxxxxxx xxxxxxxxx xxxxxxxxx xxxxxxxxx xxxxxxxxx xxxxxxxxx

@@

\end{acknowledgments}
\begin{thebibliography}{dummy}
%\bibliography
%\bibliographystyle{aasjournal}
\bibitem[Cordero et al.(2022)]{hyperrank} Cordero, J.~P., Harrison, I., Rollins, R.~P., et al.\ 2022, \mnras, 511, 2170. doi:10.1093/mnras/stac147

\bibitem[Hadzhiyska et al.(2020)]{hans} Hadzhiyska, B., Alonso, D., Nicola, A., et al.\ 2020, \jcap, 2020, 056. doi:10.1088/1475-7516/2020/10/056
\end{thebibliography}
